$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
Alignment in the example syntax

% Constructors
\newcommand{\PO}{\constr{\mathbb{0}}}
\newcommand{\comp}[2]{#1 \; \constr{\parallel} \; #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \;}
\newcommand{\send}[2]{#1 \; \constr{\langle} \; #2 \;\constr{\rangle} \;}
\newcommand{\sendp}[2]{#1 \; \constr{\langle} \; #2 \; \constr{\rangle}. \;}
\newcommand{\recv}[1]{#1 \; \constr{\mathbb{()}} \;}
\newcommand{\recvp}[2]{#1 \; \constr{(} \; #2 \; \constr{)}.\; }
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\unit}{\constr{\mathbbm{1}}}
\newcommand{\base}[1]{\constr{B[} \; #1 \; \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \; #1 \; \constr{;} \; #2 \; \constr{]}}
\newcommand{\comma}{\; \constr{,} \;}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
\todo{rename gradual to something else...}
The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first order values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

\todo{dots}
Let $x, y,\ldots$ range over \emph{variables} (including channel names) and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ are \emph{binders}, they are the only constructs which introduce bound names --- $x$ and $y$ in $P$, respectively.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \; \sendp{y}{a} \; \PO} {(\new{} y) && (\sendp{x}{y} \; \recvp{y}{z} && \; \PO)})
      && \text{names}
      \\
      & \new{} && (\comp {\recv{0} && \; \send{0}{a} \; \PO} {\new{} && (\send{1}{0} \; \recv{0} && \; \PO)})
      && \text{de Bruijn indices}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated thanks to the presence of a leftover context.
\end{itemize}

We introduce two judgments: one on variable references (\autoref{def:varref}), one on processes (\autoref{def:types}).
Both are indexed by a $\PreCtx_n$, an input $\Ctx_{idxs}$, and an output $\Ctx_{idxs}$ (the leftovers).

\begin{nidefinition}
  \label{def:varref}

  The \textsc{VarRef} relation is presented in \autoref{fig:variable-references}.
  The judgment $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ reads as ``$\gamma$ at position $i$ is of type $t$, and subtracting $y$ at position $i$ from $\Gamma$ results in $\Delta$''.

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$, which are the leftovers.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{functionality} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts.

  \begin{figure}[h]
  \begin{mathpar}
    \mprset{sep=0.5em}

@@

    \inferrule
    {loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\suc \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable reference judgments.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

\begin{example}
  Let us illustrate the use of channels with an example.
  \begin{lstlisting}[label=lst:channels,mathescape,caption=The underscore \_ introduces an anonymous declaration immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \; \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\suc \; \constr{0}$
  \end{lstlisting}
  In \autoref{lst:channels}, the constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that we can split its input annotation $\lio$, taking away $\li$ and leaving leftovers $\lo$.
\end{example}

\begin{nidefinition}
  \label{def:types}

  The typing relation \textsc{Types} is presented in \autoref{fig:types}.
  The typing judgement $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

@@

  Some comments on the leftover type system follow.

  The inaction process does not change usage annotations.

  Scope restriction expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ that is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that these are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

@@

\begin{example}
  \label{example-derivations}
  Let us type a derivation of the process $\func{p}$ using linear and shared types. 
  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transiting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process is typed following its syntax.
  On the right-hand, we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmited.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation.

>>>>

% Constructors
\newcommand{\PO}{\constr{\mathbb{0}}}
\newcommand{\comp}[2]{#1 \; \constr{\parallel} \; #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \;}
\newcommand{\send}[2]{#1 \; \constr{\langle} \; #2 \;\constr{\rangle} \;}
\newcommand{\sendp}[2]{#1 \; \constr{\langle} \; #2 \; \constr{\rangle} \; . \;}
\newcommand{\recv}[1]{#1 \; \constr{\mathbb{()}} \;}
\newcommand{\recvp}[2]{#1 \; \constr{(} \; #2 \; \constr{)} \; . \; }
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\unit}{\constr{\mathbbm{1}}}
\newcommand{\base}[1]{\constr{B[} \; #1 \; \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \; #1 \; \constr{;} \; #2 \; \constr{]}}
\newcommand{\comma}{\; \constr{,} \;}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
\todo{rename gradual to something else...}
\todo{replace judgment with judgement everywhere!}

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first order values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

Let $x, y,\ldots$ range over \emph{variables} (including channel names) and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ are \emph{binders}, they are the only constructs which introduce bound names --- $x$ and $y$ in $P$, respectively.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \sendp{y}{a} && \PO} {(\new{} y) && (\sendp{x}{y} && \recvp{y}{z} && \PO)})
      && \text{names}
      \\
      & \new{} && (\comp {\recv{0} && \send{0}{a} && \PO} {\new{} && (\send{1}{0} && \recv{0} && \PO)})
      && \text{de Bruijn indices}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context $\PreCtx_n$ and (input) usage context $\Ctx_{idxs}$, adds an extra {output usage context} to the typing rules.

This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgement with both an input and an output usage context.
  \item
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated thanks to the presence of a leftover context.
\end{itemize}
  \todo{I still don't get the last point above, as weakening and strengthening were already possible to state in \picalc{} before. What do you mean with can be stated here?}

Our type system with leftovers is composed of two typing relations: one for variable references (\autoref{def:varref}) and one for processes (\autoref{def:types}).
Both relations are indexed by a typing context $\gamma$, an input usage context $\Gamma$, and an output usage context $\Delta$ (containing the leftovers).

The typing judgement for variables $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ asserts that ``index $i$ is of type $t$ in $\gamma$, and subtracting $y$ at position $i$ from $\Gamma$ results in $\Delta$''.
The typing judgement for processes $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

\begin{nidefinition}[\textsc{VarRef}: typing variable references]
  \label{def:varref}

  The \textsc{VarRef} typing relation for variable references is presented in \autoref{fig:variable-references}.
  \begin{figure}[h]
  \begin{mathpar}
    \mprset{sep=0.5em}

@@

    \inferrule
    {loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\suc \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Typing rules for variable references.}
  \label{fig:variable-references}
  \end{figure}

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$, which are the leftovers.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{functionality} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{nidefinition}

\begin{example}[Channels]
  Let us illustrate the use of channels with an example.
  \begin{lstlisting}[label=lst:channels,mathescape,caption=The underscore \_ introduces an anonymous declaration immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \; \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\suc \; \constr{0}$
  \end{lstlisting}
  In \autoref{lst:channels}, the constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that we can split its input annotation $\lio$, taking away $\li$ and leaving leftovers $\lo$.
\end{example}

\begin{nidefinition}[\textsc{Types}: typing processes]
  \label{def:types}

  The \textsc{Types} typing relation for \picalc{} processes is presented in \autoref{fig:types}.

@@

  The inaction process does not change usage annotations.
  Scope restriction expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ that is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that these are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

@@

\begin{example}[Typing derivation]
  \label{example-derivations}
  Let us type a derivation of the process $\func{p}$ using linear and shared types. 
  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transiting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process is typed following its syntax.
  On the right-hand, we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmited.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation.
